Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-07-07 MS/ms.Rmd
Few more fixes to writing, and dealing with some more of Shin' comments before running off to meeting

Despite its widespread adoption and well established methodological procedures, meta-analysis is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widespread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. To the uninitiated, this concern might not be unfounded given that heterogenity in effects is often high in ecological and evolutionary meta-analyses [@Senior2016], telling us that the effect size varies a great deal from one population to another. However, to many, the very fact that highly heterogeneous effects exist is the core strength of meta-analysis when considered carefully [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. Indeed, exploring the reasons for why effects are likely to vary across studies "...is the spice of life" [p.g., 519; @Cooperetal2019].

@@

Meta-analysis is always concerned with effect heterogeneity. In other words, the factors that drive differences in the direction and magnitude an of effect within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019]. The concept of heterogeneity is vitally important because it tells us how general a specific conclusion is likely to be, how much of the variance we see is the result of real biology (and not just sampling variance), and provides a way to understand what the key biological factors are leading to an effect. There are a plethora of phenomena driving effect heterogeneity, some of which are interesting and others that are less so. Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability in an effect because of the inherent variability in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating this 'true' effect. In meta-analysis we formally include sampling variance to remove this source of heterogeneity directly - hence why having sampling variances for an effect is so important. Second, there are sociological factors impacting heterogeneity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions on what we hypothesize are key players impacting an effect improving our understanding of the biological world around us. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it's important we know and understand this as it can determine how experiments are designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' - factors we know vary, but are rather unsurprising to us. These can include differences in temperature, pH, dosages, water potential etc. that are obviously going to vary and have an impact on the magnitude of an effect. In some cases, these factors might actually not be 'nuisance' variables but rather the main interest. However, in many cases the can result in a more complicated interpretation of an effect measure then what is desired.

All the various sources of heterogeneity (except sampling variance) all contribute (to varying extents) to the 'apples and oranges' problem often touted as being a problem in meta-analyses (or quantitative syntheses more generally). Combining standardised effects, while ignoring these sources of heterogeneity can result in the meaning of a meta-analysis not being particularly useful to many. Many others, however, argue that this is just what we are interested in synthesizing. In the words of Gene Glass, the father of meta-analysis: "Of course it mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest and the population one wishes to make conclusions about. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. <!---  as said, comparative analysis by definition, compares applies and oranges - = different species - I guess you are mainly talking about mixing different methods ones - this is where I said earlier - focusing on explaining hetero comes in. Also, mixing different quality of studies are often referred to as "Garbage in and garbage out" ---> Instead, if we are interested in generalising only to apples then mixing fruit will not be ideal. We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms then including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are known to differ. By synthesizing diverse study outcomes, it allows us to not just report on overall effects (if the question is relevant to ask), but understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis - an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. <!--- this is know as putting one number on so many different kinds of studies - but of course, this may not be that interesting apart from a starting point for subsequent meta-regression analyses ---> For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may be misleading. It is important to consider these questions when conducting meta-analyses.

Of course, there are certain types of heterogeneity that transcend different questions. In other words, if we compare different types of 'apples' is it still fair to synthesize these studies to understand overall effects? This situation might be analogous to a meta-analysis in comparative physiology that attempts to understand the effects of temperature on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of fish under different temperatures, but in the end, they are still all the same fish species, governed by the same thermodynamic effects on growth. In this situation, we might still be interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. It is not always clear how we should tackle these sources of heterogeneity. If temperature differences are of direct interest, then we should definitely model them. If, however, it is less interesting, than say the habitat occupied then it may introduce an unnecessary complication to the interpretation of the effect(s) of interest. Regardless, a diversity of factors *a priori* that are hypothesised or known to impact heterogeneity does not preclude us from also exploring all these important moderators or predictors of variability in effects across studies; however, depending on the effect size used it may just complicate their interpretation or make modelling of all these effects challenging.

>>>>

Despite its widespread adoption and well established methodological procedures, meta-analysis is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widespread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. To the uninitiated, this concern might not be unfounded given that heterogenity in effects is often high in ecological and evolutionary meta-analyses [@Senior2016], telling us that the effect size varies a great deal from one population to another. However, to many, the very fact that highly heterogeneous effects exist is a core strength of meta-analysis [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. Indeed, exploring the reasons for why effects are likely to vary across studies "...is the spice of life" [p.g., 519; @Cooperetal2019].

@@

Meta-analysis is always concerned with effect heterogeneity; in other words, the factors that drive differences in the direction and magnitude an of effect within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019]. The concept of heterogeneity is vitally important because it tells us how general a specific conclusion is likely to be, how much of the variance we see is the result of real biology (and not just sampling variance), and provides a way to understand what the key biological factors are leading to an effect [@Nakagawa2017; @Borenstein2019]. There are a plethora of phenomena driving effect heterogeneity, some of which are interesting and others that are less so.

Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability driven by the inherent uncertainty in sampling of a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating this 'true' effect. In meta-analysis we formally include sampling variance to remove this source of heterogeneity directly - hence why having sampling variances for an effect is so important [@Gurevitch2018; @Koricheva2013]. Second, there are sociological factors impacting heterogeneity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions from hypotheses we believe are key players impacting an effect. This improves our understanding of the biological world around us and helps inform on future research directions. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it is important we know and understand this as it can determine how experiments are designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' [@Cooperetal2019] - factors we know vary, but are rather unsurprising to us. These can include differences in temperature, pH, dosages, water potential etc. that are obviously going to vary and have an impact on the magnitude of an effect. In some cases, these factors might actually not be 'nuisance' variables but rather the main interest of a synthesis. However, in many cases they can result in a more complicated interpretation of an effect measure then what is desired.

All the various sources of heterogeneity (except sampling variance) all contribute (to varying extents) to the 'apples and oranges' problem often touted as being an issues with some meta-analyses (or quantitative syntheses more generally). Indeed, some of this criticism revolves around combining studies that vary in their quality (e.g., observational vs experimental). Combining standardised effects, while ignoring these sources of heterogeneity can result in the meaning of a meta-analysis not being particularly useful to many. However, many argue that this is just what we are interested in synthesizing [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. In the words of Gene Glass, the father of meta-analysis: "Of course it mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest, the population one wishes to make conclusions about and how hetergenous effects actually are. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. Instead, if we are interested in generalising only to apples then mixing fruit will not be ideal. In addition, if we want to understand why different 'fruit' vary in their effects, then we can only ask this question by including all fruit [@Borenstein2019].

We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms then including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are known to differ. By synthesizing diverse study outcomes, it allows us to not just report on overall effects (if the question is relevant to ask), but understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis, and how this can affect interpretations made - an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincies with low heterogenity esatimtes, then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogenity, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may be misleading. It is important to consider these questions when conducting meta-analyses.

Of course, there are certain types of heterogeneity that transcend different questions or that are simply a 'nuisance' to interpreting effects. In other words, if we compare different types of 'apples' is it still fair to synthesize these studies to understand overall effects? This situation might be analogous to a meta-analysis in comparative physiology that attempts to understand the effects of temperature on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of fish under different temperatures, but in the end, they are still all the same fish species, governed by the same thermodynamic effects on growth. In this situation, we might still be interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. It is not always clear how we should tackle these sources of heterogeneity. If temperature differences are of direct interest, then we should definitely model them. If, however, it is less interesting, than say the habitat occupied, then it may introduce an unnecessary complication to the interpretation of the effect(s) of interest. Regardless, a diversity of factors *a priori* that are hypothesised or known to impact heterogeneity does not preclude us from also exploring all these important moderators or predictors of variability in effects across studies; however, depending on the effect size used it may just complicate their interpretation or make modelling of all these effects challenging.
